### 2015-07-15 Harris_2015/manuscript-materials/draft.Rmd
more

***Introduction***

If nontrophic species interactions, such as competition, are important drivers
of community assembly, then ecologists might expect to see their influence in
our data sets [@macarthur_population_1958; @diamond_island_1975]. Despite
decades of work and several major controversies, however [@lewin_santa_1983;
@strong_ecological_1984; @gotelli_swap_2003; @connor_checkered_2013], existing
methods for detecting competition's effects on community structure are
unreliable [@gotelli_empirical_2009], and thus important ecological processes
remain poorly understood. More generally, it can be difficult to reason about
the complex web of direct and indirect species interactions in real
assemblages, especially when these interactions occur against a background of
other ecological processes such as dispersal limitation and environmental filtering
[@connor_checkered_2013]. For this reason, it isn't always clear what kinds of
patterns would even constitute evidence of competition, as opposed to some
other biological process or random sampling error
[@roughgarden_competition_1983; @lewin_santa_1983].

Most existing methods in this field compare the frequency with which two
putative competitors are observed to co-occur, versus the frequency that would
be expected if *all* species on the landscape were independent
[@strong_ecological_1984; @gotelli_empirical_2009]. Examining a species pair
against such a "null" background, however, rules out the possibility that the
overall association between two species could be driven by an outside force.
For example, even though the two shrub species in Figure 1 compete with one
another for resources at a mechanistic level, they end up clustering together
on the landscape because they both grow best in areas that are not
overshadowed by trees. If this sort of effect is common, then significant
deviations from independence will not generally provide convincing
evidence of species' direct effects on one another.

While competition between the two shrubs in the previous example does not
leave the commonly-expected pattern in community structure (negative
association at the landscape level), it nevertheless does leave a signal in
the data (Figure 1C). Specifically, *among shaded sites*, there will be a
deficit of co-occurrences, and *among unshaded sites*, there will also be such
a deficit.

In this paper, I introduce Markov networks [undirected graphical models also
known as Markov random fields; @murphy_machine_2012] as a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them with observational data. Markov networks,
which generalize partial correlations to non-Gaussian data
[@lee_learning_2012; @loh_structure_2013], have been used in many scientific
fields for decades to model associations between various kinds of "particles".
For example, a well-studied network named the Ising model has played an
important role in our understanding of physics [where nearby particles tend to
align magnetically with one another; @cipra_introduction_1987]. In spatial
contexts, these models have been used to describe interactions between
adjacent grid cells [@harris_contact_1974; @gelfand_modelling_2005]. In
neurobiology, they have helped researchers determine which neurons are
connected to one another by modeling the structure in their firing patterns
[@schneidman_weak_2006]. Following recent work by @azaele_inferring_2010 and
@fort_statistical_2013, I suggest that ecologists could similarly treat
species as the interacting particles in the same modeling framework. Doing so
would allow ecologists to simulate and study the landscape-level consequences
of arbitrary species interaction matrices, even when our observations are not
Gaussian. While ecologists explored some related approaches in the 1980's
[@whittam_species_1981], computational limitations had previously imposed
severe approximations that produced unintelligible results [e.g.
"probabilities" greater than one; @gilpin_factors_1982].  Now that it is
computationally feasible to fit these models exactly, the approach has become
worth a second look.

@@

***Methods***

***Conditional relationships and Markov networks.***
Ecologists are often interested in learning about direct interactions between
species, controlling for the indirect influence of other species. In
statistical terms, this implies that ecologists want to estimate *conditional*
("all-else-equal") relationships, rather than *marginal* ("overall") relationships.
The most familiar conditional relationship
is the partial correlation, which indicates the portion of the sample correlation
between two species that remains after controlling for other variables in the
data set [@albrecht_spatial_2001]. The example with the shrubs and trees in
Figure 1 shows how the two correlation measures can have opposite signs,
and suggests that the partial correlation is more relevant
for drawing inferences about species interactions (e.g. competition).
Markov networks extend this approach to non-Gaussian data, much as
generalized linear models do for linear regression [@lee_learning_2012].

@@

***Simulations.***
In order to compare different methods, I simulated two sets of landscapes using known
parameters.

@@

***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
In the previous two sections, the values of $\alpha$ and $\beta$ were known. In practice,
however, ecologists will often need to estimate these parameters from co-occurrence data.
When the number of species is reasonably small, one can compute exact maximum
likelihood estimates for all of the $\alpha$ and $\beta$ coefficients by optimizing
$p(\vec{y}; \alpha, \beta)$. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
[@murphy_machine_2012], ensuring that this procedure will always converge on
the global maximum. This maximum is the unique combination of $\alpha$ and $\beta$
coefficients that would be expected to produce exactly the observed co-occurrence
frequencies. For the analyses in this paper, I used the *rosalia* package
[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to
define the objective function and gradient as R code.  The rosalia package then
uses the `BFGS` method in R's `optim` function to find the best values for $\alpha$
and $\beta$.

@@

***Comparison with existing techniques.***
After fitting Markov networks to the simulated landscapes described above, I
used several other techniques for determining the sign and strength of marginal
associations between pairs of species (Appendix B).

@@

***Method evaluation.***
For the simulated landscapes based on Figure 1, method evaluation was fairly
qualitative: any method whose test statistic for the two shrubs indicated a
negative relationship passed; other methods failed.

@@

***Results***

***Three species.***
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---due
to indirect effects of the dominant tree species. As a result, the
covariance method falsely reported positive associations for 94% of the simulated
landscapes, and the randomization-based null model falsely reported such
associations 100% of the time. The two methods for evaluating conditional
relationships (Markov networks and partial covariances), however, successfully
controlled for the indirect pathway via the tree species and each correctly
identified the direct negative interaction between the shrubs 94% of the time.

***Larger landscapes.***
The accuracy of the four evaluated methods varied substantially, depending on
the parameters that produced the simulated communities (Figure 3). In general,
however, there was a consistent ordering: overall, the Markov network
explained 54% of the "true" parameters' squared deviations from zero, followed
by partial covariances (33%), and sample covariances (22%). The null model
scores initially explained only 12%. After manually reducing the value of one
especially strong outlier ($Z=1004$, implying $p<10^{-1000000}$), this
increased to 17% (Appendix B).  Figure 3 reflects the adjusted version of the
results.

***Discussion***

The results presented above show that Markov networks can reliably recover
species' pairwise interactions from observational data, even for cases where
a common null modeling technique reliably fails. Specifically, Markov networks
were successful even when direct interactions were largely overwhelmed by indirect
effects (Figure 1) or environmental effects (lower panels of Figure 3). For cases
where fitting a Markov network is computationally infeasible, these results also
indicate that partial covariances---which can be computed straightforwardly by linear
regression---can often provide a surprisingly useful approximation. The
partial correlations' success on simulated data may not
carry over to real data sets, however; @loh_structure_2013 show that the linear
approximations can be less reliable in cases where the true interaction matrix
contains more structure (e.g. guilds or trophic levels). On the other hand,
if ecologists are familiar enough with the natural history of their study
systems to describe this kind of structure as a prior distribution on the
parameters or as a penalty on the likelihood, then this information could
reduce the effective degres of freedom to estimate and real-world results might
be even better than those shown in Figure 3.

@@

***Acknowledgements:***
This research was funded by a Graduate Research Fellowship from the US
National Science Foundation and benefited greatly from discussions with A.
Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor.
Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
Golding, and N. J. Gotelli provided useful feedback on an earlier draft of
this work.

>>>>

\setlength{\parindent}{1cm}

\noindent  
***Introduction***

If nontrophic species interactions, such as competition, are important drivers
of community assembly, then ecologists might expect to see their influence in
our data sets [@macarthur_population_1958; @diamond_island_1975]. Despite
decades of work and several major controversies, however [@lewin_santa_1983;
@strong_ecological_1984; @gotelli_swap_2003; @connor_checkered_2013], existing
methods for detecting competition's effects on community structure are
unreliable [@gotelli_empirical_2009]. In particular, species' effects on one
another can become lost in the complex web of direct and indirect interactions
in real assemblages. For example, the competitive interaction between the
two shrub species in Figure 1A can become obscured by their shared tendency
to occur in unshaded areas (Figure 1B). If this sort of effect is common, then
the vast majority of ecologists' methods, which rely on test statistics describing
the total number of co-occurrences or overall correlation between putatively
interacting species [@diamond_island_1975; @strong_ecological_1984;
@gotelli_empirical_2009; @veech_probabilistic_2013; @pollock_understanding_2014]
will not generally provide much evidence regarding species' direct effects on
one another.

While competition doesn't reliably leave the expected pattern in community
structure (a low co-occurrence rate at the landscape level), it nevertheless does
leave a signal in the data (Figure 1C). Specifically, *among shaded sites*,
there will be a deficit of co-occurrences between the competing shrub species,
and *among unshaded sites*, there will also be such a deficit. More generally,
we can obtain much better estimates of the association between two species from
their conditional relationships (i.e. by controlling for other species in the
network) than we could get from their overall co-occurrence rates.

In this paper, following recent work by @azaele_inferring_2010 and
@fort_statistical_2013, I show that Markov networks [undirected graphical models
also known as Markov random fields; @murphy_machine_2012] can provide a framework
for understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observational data. Markov networks
have been used in many scientific fields for decades to model associations
between various kinds of "particles". For example, a well-studied network named the Ising model has played an
important role in our understanding of physics [where nearby particles tend to
align magnetically with one another; @cipra_introduction_1987]. In spatial
contexts, these models have been used to describe interactions between
adjacent grid cells [@harris_contact_1974; @gelfand_modelling_2005]. In
neurobiology, they have helped researchers determine which neurons are
connected to one another by modeling the structure in their firing patterns
[@schneidman_weak_2006].

Doing so
would allow ecologists to simulate and study the landscape-level consequences
of arbitrary species interaction matrices, even when our observations are not
Gaussian. While ecologists explored some related approaches in the 1980's
[@whittam_species_1981], computational limitations had previously imposed
severe approximations that produced unintelligible results [e.g.
"probabilities" greater than one; @gilpin_factors_1982].  Now that it is
computationally feasible to fit these models exactly, the approach has become
worth a second look.

@@

\noindent  ***Methods***

\noindent  
***Conditional relationships and Markov networks.***
Ecologists are often interested in learning about direct interactions between
species, controlling for the indirect influence of other species. In
statistical terms, this implies that ecologists want to estimate *conditional*
("all-else-equal") relationships, rather than *marginal* ("overall") relationships.
The most familiar conditional relationship
is the partial correlation, which indicates the portion of the sample correlation
between two species that remains after controlling for other variables in the
data set [@albrecht_spatial_2001]. The example with the shrubs and trees in
Figure 1 shows how the two correlation measures can have opposite signs,
and suggests that the partial correlation is more relevant
for drawing inferences about species interactions (e.g. competition).
Markov networks extend this approach to non-Gaussian data, much as
generalized linear models do for linear regression [@lee_learning_2012].

@@

\noindent  
***Simulations.***
In order to compare different methods, I simulated two sets of landscapes using known
parameters.

@@

\noindent  
***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
In the previous two sections, the values of $\alpha$ and $\beta$ were known. In practice,
however, ecologists will often need to estimate these parameters from co-occurrence data.
When the number of species is reasonably small, one can compute exact maximum
likelihood estimates for all of the $\alpha$ and $\beta$ coefficients by optimizing
$p(\vec{y}; \alpha, \beta)$. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
[@murphy_machine_2012], ensuring that this procedure will always converge on
the global maximum. This maximum is the unique combination of $\alpha$ and $\beta$
coefficients that would be expected to produce exactly the observed co-occurrence
frequencies. For the analyses in this paper, I used the *rosalia* package
[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to
define the objective function and gradient as R code.  The rosalia package then
uses the `BFGS` method in R's `optim` function to find the best values for $\alpha$
and $\beta$.

@@

\noindent  
***Comparison with existing techniques.***
After fitting Markov networks to the simulated landscapes described above, I
used several other techniques for determining the sign and strength of marginal
associations between pairs of species (Appendix B).

@@

\noindent  
***Method evaluation.***
For the simulated landscapes based on Figure 1, method evaluation was fairly
qualitative: any method whose test statistic for the two shrubs indicated a
negative relationship passed; other methods failed.

@@

\noindent  
***Results***

\noindent  
***Three species.***
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---due
to indirect effects of the dominant tree species. As a result, the
covariance method falsely reported positive associations for 94% of the simulated
landscapes, and the randomization-based null model falsely reported such
associations 100% of the time. The two methods for evaluating conditional
relationships (Markov networks and partial covariances), however, successfully
controlled for the indirect pathway via the tree species and each correctly
identified the direct negative interaction between the shrubs 94% of the time.

\noindent  
***Larger landscapes.***
The accuracy of the four evaluated methods varied substantially, depending on
the parameters that produced the simulated communities (Figure 3). In general,
however, there was a consistent ordering: overall, the Markov network
explained 54% of the "true" parameters' squared deviations from zero, followed
by partial covariances (33%), and sample covariances (22%). The null model
scores initially explained only 12%. After manually reducing the value of one
especially strong outlier ($Z=1004$, implying $p<10^{-1000000}$), this
increased to 17% (Appendix B).  Figure 3 reflects the adjusted version of the
results.

\noindent  
***Discussion***

\noindent  
The results presented above show that Markov networks can reliably recover
species' pairwise interactions from observational data, even for cases where
a common null modeling technique reliably fails. Specifically, Markov networks
were successful even when direct interactions were largely overwhelmed by indirect
effects (Figure 1) or environmental effects (lower panels of Figure 3). For cases
where fitting a Markov network is computationally infeasible, these results also
indicate that partial covariances---which can be computed straightforwardly by linear
regression---can often provide a surprisingly useful approximation. The
partial correlations' success on simulated data may not
carry over to real data sets, however; @loh_structure_2013 show that the linear
approximations can be less reliable in cases where the true interaction matrix
contains more structure (e.g. guilds or trophic levels). On the other hand,
if ecologists are familiar enough with the natural history of their study
systems to describe this kind of structure as a prior distribution on the
parameters or as a penalty on the likelihood, then this information could
reduce the effective degres of freedom to estimate and real-world results might
be even better than those shown in Figure 3.

@@

\noindent  
***Acknowledgements:***
This research was funded by a Graduate Research Fellowship from the US
National Science Foundation and benefited greatly from discussions with A.
Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor.
Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
Golding, and N. J. Gotelli provided useful feedback on an earlier draft of
this work.

\setlength{\parindent}{0cm}
